NFL Draft Analysis

### 2020-05-11 draft_analysis.Rmd
Some reformatting and updated prose

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = -1)
library(tidyverse)
library(rvest)
library(gridExtra)
setwd("/Users/PhilipIpe/Documents/Data Science/Projects/NFL Draft Analytics/")
```

@@

@@

We will also scrape the data from the 2020 draft for use in our machine learning problem and save it to our df variable.

@@

Load the NFL draft history data, containing the draft data from rounds 1-7 from the 1970-2016 NFL draft. Remove all of the missing data entries, and convert the necessary columns to integers. 
```{r cars}
head(master_df)

@@

master_df$Age <- as.integer(master_df$Age)
master_df$CarAV <- as.integer(master_df$CarAV)
master_df$DrAV <- as.integer(master_df$DrAV)
master_df$G <- as.integer(master_df$G)
master_df$Pick <- as.integer(master_df$Pick)
```

Now let's process our 2020 Draft data. First, we will get rid of columns we don't care about, then remove blank rows, add year drafted to each observation, and convert the necessary columns to integers, all like we did for the historical draft dataset. We will also filter the data to only include quarterbacks, wide receivers, and running backs.

@@

df_2020$Pick <- as.integer(df_2020$Pick)
df_2020$Age <- as.integer(df_2020$Age)
df_2020$Tm <- as.factor(df_2020$Tm)
df_2020$Pos <- as.factor(df_2020$Pos)
```

Now let's create a dataframe in which we can compare different positions that were drafted. First, we want to break up our data into 4 periods. Then for each period, we will take average career value of each position grouped by each round that the player was selected. We will make a new pos_df dataframe with these groupings.

@@

head(pos_df)
```

Let's also add a dataframe that contains a variable that is TRUE when the player that was drafted was Offensive Rookie of the Year, and FALSE otherwise. We will also filter out all of the defensive players in this new dataframe.

@@

off_df$roty <- as.factor(off_df$roty)
```
# Part 3: Exploratory Data Analysis

## Trends in Average value
Has average value for players selected in different rounds gone up over time?
Have GMs gotten better at drafting in the first round, second round, etc...?

@@

## Analyzing the Quarterback position

@@

## Analyzing the WR and RB positions

@@

Now, right off the bat, we hope that the model will predict that Joe Burrow has a higher likelihood of winning the MVP. Not only because he's a stud, but because he's fresh off the Heisman, and the first overall pick QB that is going to a team in desperate need of good QB play. Vegas has him as the clear favorite to win, and the season is set up for him to take the title, so if our model is worth anything

First, lets control the resampling of our data. In this case, we’re going to cross-validate the data 3 times, therefore training it 3 times on different portions of the data before settling on the best tuning parameters (for gbm it is trees, shrinkage, and interaction depth).

@@

library(caret)

@@

# Creating our Model
Now to actually train the model...

@@

@@

The summaries tell us how much influence each variable had on the model. It was interesting to see that position had such absolutely zero influence in predicting the model. This tells us that quarterbacks, wide receivers, or running backs do not have a leg up on each other for the MVP award!

What is even more interesting is that the team the player was drafted to seemed to influence the model more than the pick number itself! Before I had trained the model, I expected pick, position, and then team to have relative influence in that order!

@@

predictions <- predict(object=model, df_2020[,predictor_names], type='prob')
head(predictions)
```

@@

Now let's get the top 7 players that the model is most certain will be the Offensive Rookie of the Year

@@

top10_roty
```

Joe Burrow is our model's top prediction to win the MVP!!! We also see that Cam Akers, Tua, Henry Ruggs III, Jerry Jeudy, CeDee Lamb, and Justin Jefferson also appear in the top 10. All of these players currently have the highest odds to win the the Offensive MVP according to OddsShark.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = -1)
library(tidyverse)
library(rvest)
library(gridExtra)
library(caret)
library(knitr)
setwd("/Users/PhilipIpe/Documents/Data Science/Projects/NFL Draft Analytics/")
```

@@

The result will be stored in the dataframe *master_df* with entities for each player.

Note: We will also do some process in this step as well, but we will go into more detail on that when we repeat the processing steps in another dataframe later in the walkthrough.

@@

We will also scrape the data from the 2020 draft in the same way for use in our machine learning problem and save it to our *df_2020* variable.

@@

Load the NFL draft history data from *master_df*, containing the draft data from rounds 1-7 from the 1970-2016 NFL draft. Remove all of the missing data entries, and convert the necessary columns to integers.

```{r process_master}

@@

master_df$Age <- as.integer(master_df$Age)
master_df$CarAV <- as.integer(master_df$CarAV)
master_df$DrAV <- as.integer(master_df$DrAV)
master_df$G <- as.integer(master_df$G)
master_df$Pick <- as.integer(master_df$Pick)

kable(head(master_df))
```

Our *master_df* dataframe now has entities for each player with the following variables:

    Rnd : Round in which the player was selected

    Pick: Overall selection in draft

    Tm  : Team

    Player: Player name

    Pos : Position

    Age

    To: Last year in the league

    AP1: First team All-Pro selections

    PB: Pro Bowl selections

    St: Number of years as primary starter for his team at his position

    CarAv: Weighted Career Approximate Value

                  = 100% of AV of his best season + 95% of AV of his next best season + 90% of AV of his third-best season + ...

    DrAV: Average value accumulated for the team that drafted this player

    G : Games Played

    year: Year drafted

Now let's process our 2020 Draft data. First, we will get rid of columns we don't care about, then remove blank rows, add year drafted to each observation, and convert the necessary columns to integers, all like we did for the historical draft dataset. We will also filter this dataframe to only include quarterbacks, wide receivers, and running backs.

@@

df_2020$Pick <- as.integer(df_2020$Pick)
df_2020$Age <- as.integer(df_2020$Age)
df_2020$Tm <- as.factor(df_2020$Tm)
df_2020$Pos <- as.factor(df_2020$Pos)

kable(head(df_2020))
```

So, now our *df_2020* dataframe contains only the offensive skill position players that were drafted in 2020 as entities. The variables of this dataframe are **Rnd**, **Pick**, **Tm**, **Player**, **Pos**, **Age**, and **Year** like we have in our *master_df* variable.

Now let's create a dataframe in which we can compare different positions that were drafted. First, we want to break up our data into 4 periods. Then for each period, we will take average career value of each position grouped by each round that the player was selected. We will make a new *pos_df* dataframe with these groupings.

@@

kable(head(pos_df))
```

Let's also add a dataframe *off_df* that is the same as *master_df*, but contains a variable **roty** that is Yes when the player that was drafted from 1970-2016 was Offensive Rookie of the Year, and No otherwise. We will also filter the dataframe to only contain offensive skill position players, like we did for the 2020 draft data.

@@

off_df$roty <- as.factor(off_df$roty)

kable(head(off_df))
```

# Part 3: Exploratory Data Analysis

## Trends in Average Value

Has average value for players selected in different rounds gone up over time?
Have GMs gotten better at drafting in the first round, second round, etc...?

@@

## Analyzing the Quarterback Position

@@

## Analyzing the WR and RB

@@

Now, right off the bat, we hope that the model will predict that Joe Burrow has a higher likelihood of winning the MVP. Not only because he's a stud, but because he's fresh off the Heisman, and the first overall pick QB that is going to a team in desperate need of good QB play. Vegas has him as the clear favorite to win, and the season is set up for him to take the title, so if our model is worth anything, it should at the very least have Joe Burrow as one of its top predictions for the award.

First, lets control the resampling of our data. In this case, we’re going to cross-validate the data 3 times, therefore training the model 3 times on different portions of the data before settling on the best tuning parameters. For gbm, the tuning parameters are trees, shrinkage, and interaction depth.

@@

@@

# Creating our Model

Now to actually train the model...

First, we will set the predictor variables to **Tm**, **Pos**, and **Pick**, and the sole outcome variable as **roty**. Then we will use the *train()* method from the caret package to train the model using the *off_df* variable that contains all our historical draft data on all the offensive skill position players.

@@

We will print the summary of the model below, which will display the relative influences of each of the predictor variables.

@@

The summaries tell us how much relative influence each variable had on the model. It was interesting to see that position had such absolutely zero influence in predicting the model. This tells us that quarterbacks, wide receivers, or running backs do not have a leg up on each other for the MVP award!

What is even more interesting is that the team the player was drafted to seemed to influence the model more than the pick number itself! Before I had trained the model, I expected pick, position, and then team to have relative influence in that order.

@@

predictions <- predict(object=model, df_2020[,predictor_names], type='prob')
kable(head(predictions))
```

@@

Now let's get the top 10 players that the model is most certain will be the Offensive Rookie of the Year

@@

kable(top10_roty)
```

Joe Burrow is our model's top prediction to win the MVP!!! We also see that Cam Akers, Tua, Henry Ruggs III, Jerry Jeudy, CeDee Lamb, and Justin Jefferson also appear in the top 10. All of these players are among those that have the highest odds to win the the Offensive MVP according to OddsShark.

Again, it's important to highlight that we trained this model JUST based on the pick number, team, and position for each player, whereas the Vegas odds have most certainly taken into account the player's performance in college as well.
